Dusty Days and a Vibrio Craze: The Influence of Saharan Dust Deposition on Microbial Communities in High Nutrient Coastal Waters 

### 2023-05-03 products/manuscript/Manuscript.qmd
Grammer /citations checks for manuscript

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year [@dalmeidaModelSaharanDust1986\]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico. These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria [@kelloggCharacterizationAerosolizedBacteria2004\], fungi [@ramirez-camejoFungalDiversitySahara2022\], virus-like particles [@griffinAfricanDesertDust\], and minerals [@formentiChemicalCompositionMineral2003\]. These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron [@millsIronPhosphorusColimit2004; @savoieNonseasaltSulfateNitrate1989; @grahamAtmosphericTransportPhosphorus1982\] . This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups [@lenesSaharanDustPhosphatic2008; @westrichSaharanDustNutrients2016a ; @westrichSaharanDustNutrients2016a].

@@

Previous work has characterized *Vibrio* blooms in response to Saharan dust input in the oligotrophic setting of the Caribbean and subtropical Atlantic (Barbados and Florida Keys, respectively), with surface water concentrations of total *Vibrio* increasing by five to thirty times that found during non-dust conditions and returning to baseline levels within 24-48 hours. *Vibrio* composition within the larger microbial community also shifted following dust deposition, with initial levels of \<1.4% to a peak of 19.8% of the bacterial community [@westrichSaharanDustNutrients2016a]. A similar phenomenon was also observed in the surface waters of the tropical and open ocean mid-Atlantic with *Vibrio* populations increasing 1.5-fold in the mid-Atlantic following deposition [@westrichVibrioPopulationDynamics2018a] In a follow up study in the Florida Keys, episodic dust events during a daily time series promoted a succession of bacterial responses, with declines in *Prochlorococcus* coinciding with initial increases in bacteria belonging to the order *Vibrionales*, followed by subsequent shifts in response of different bacterial groups [@borchardtSaharanDustDeposition2020b]. To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients. However, the composition of desert dust can be complex and may have the potential to deliver critical resources that could be exploited by opportunistic microbes (like *Vibrio*) and elicit a growth response despite higher baseline nutrient levels. Given that human exposure is likely in these coastal waters, it is important to elucidate the relationship between dust input and microbial blooms in these higher nutrient environments.

@@

Sampling took place at three locations in Corpus Christi, TX: Blind Oso Bay, a residential canal system on Padre Island, and the Gulf. Sites were chosen based on differing baseline nutrient levels (defined here as the 2022 average from monthly observations). Blind Oso Bay is a shallow tributary that is popular amongst wadefisherman and kayakers. The Bay is often impacted by fluctuating salinities and high levels of nutrients (baseline:13.03 µM nitrate+nitrite, 4.44 µM orthophosphate) and chlorophyll due to its proximity to wastewater treatment discharge (Wetz, 2014).­ Additionally, the Bay has persistent issues with high levels of fecal indicator bacteria, placing it on the U.S. impaired waters list (Nicolau & Hill, 2014, Texas Commission on Environmental Quality, 2022). The residential canal system (baseline: 0.69µM nitrate+nitrite, 0.20 µM orthophosphate), and can also be impacted by high nutrients and chlorophyll from storm water runoff; salinity is primarily driven by precipitation. The Gulf site is characterized by constant salinities (\~36) and relatively low baseline nutrient levels (baseline: 0.19µM nitrate+nitrite and 0.36µM orthophosphate).

@@

This project works with three primary data sets. (1) Enumeration of Vibrio bacteria as copies/mL from qPCR (copies_master.csv), (2) Dust concentrations as aerosol optical thickness (AOT) derived from the Naval Research Lab (nrl_conc.csv), and (3) Temperature and Salinity across the daily time series (ysi_2.csv). Below is a brief summary of how each data set was cleaned. More details can be found in the supplementary files, containing the code and comments describing what each line does. Please refer to the `project_README.md` file located in the NathanGreenslit-MADA-Project folder for details on the repository contents and instructions on reproducibility.\

@@

Quantitative PCR provides enumeration results as "cycle threshold or Cq" values. This depicts the cycle number in which enough of the DNA target was present to be amplified and thus detected. Lower Cq values correspond to higher target concentrations (as it takes less cycles to amplify), and vice verse for higher Cq values. The cleaning of this data set consisted of converting these Cq values into something that can be used to quantify Vibrio (copies of target per mL of seawater). To do so, I needed to take into account the amount of water sample that was concentrated on a filter, the total amount of DNA that was eluted following a DNA extraction, and the amount of DNA template added to each mix for qPCR (to name a few). Taking these into account, we are able to calculate our way from Cq values to copies of Vibrio per mL of seawater. qPCR was conducted in triplicate, so the last step was to take the average of the three replicates to have a final value per sample. A date column was also added for each sample.

@@

This figure depicts the summed dust concentration 24hr prior to sample collection (time points 13hr--> 7hr the next day) across the daily time series. The daily time series captured two dust events. One light event starting on the 13th of July, where AOD reached 0.2, and a much heavier dust event from the 16th to the 18th of July with AOD = 55-57. Dust AOD returned to lower concentrations following the time series.

@@

```{r, echo=FALSE, message=FALSE, fig.cap="Figure 3: Envonmental parameters across daily time series with color by site"}
knitr::include_graphics(here("results", "plots", "env.png"))
```

@@

Previous literature has shown that Vibrio respond 12-24 hours following dust deposition events. Based on this, we need to look at lags between the two times series for dust data and growth data. To do this, we can set up a cross-correlation analysis, which allows us to examine the lag/lead relationship between the two variables.

CCF plots showed that BO has a significant lag relationship a value of -1. In other words, dust occurs, and one day later, we see a response in copies_mL. C2 has a lag at -2, and RD at -1.

@@

#### Linear Regressions

@@

Multi-variate models that included dust, water chemistry, nutrients, and DOM performed the best, with mean RMSE values much lower than the Null Model.

@@

*Vibrio* gene copies/mL was found to significantly correlate with dust aerosol optical density at Blind Oso (r = 0.7, p = 0.01), Canals (r = 0.75, p = 0.005), and the Gulf (r = 0.67, p = 0.02). The introduction of dust constituents have the potential to serve as a temporary alleviation from the limitation of nutrients that are essential for bloom formation. *Vibrio* species have one of the fastest doubling times among bacteria [@gibsonDistributionBacterialDoubling2018 ; @eagonPSEUDOMONASNATRIEGENSMARINE1962\], and are able to exploit introduced nutrients due their rapid motility, diverse genomic repertoire, and iron-chelating complexes [@ringgaardChemotaxisArraysVibrio2018; @okadaVibriosCommonlyPossess2005 ; @payneVibrioIronTransport2016]. Initially comprising a small proportion of the marine microbial communities (\<1%), *Vibrio* can come to quickly dominate, and thus have been classified as 'oppurtunitrophs'. The following declines in *Vibrio* gene copies/mL may be attributed to top down control, such as viral lysis [@molina-quirozInteractionsVibrioPhages2023], or bottom-up controls due to an exhaustion of the resources provided by the dust [@psennerAreRapidChanges1992].

Similar to previous findings set in the Florida Keys and North Atlantic [@westrichVibrioPopulationDynamics2018a; @westrichSaharanDustNutrients2016a], peaks in growth were observed 24-48 hours following dust deposition. Comparing peak *Vibrio* gene copies/mL to initial levels on the 7th of July, the Gulf experienced a 338-fold increase in *Vibrio* gene copies/mL, Blind Oso a 3.4-fold increase, and the Canals a 1.3-fold increase.

@@

The Gulf exhibited a sharper increase during the second, heavier dust event, while Blind Oso exhibited a smaller increase during the heavy dust event (Figure 2). This difference in response may be attributed to the nutrient dynamics of the sites. In marine waters, nutrients are typically limiting due to tight microbial recycling. Therefore the introduction on nutrients at such a site may result in rapid uptake and utilization. At Blind Oso, a priming effect may be occurring, where dust constituents resulting from a secondary flux may not be as essential for growth following an initial deposition event, where the constituents are reaching a maximum in how much they can support growth.

@@

While it can be proposed that the degree of microbial response is contingent on the environmental settings of the site, further work is needed to look at each specific environmental parameter and how it may impact *Vibrio* abundance. At Blind Oso for example, the drastic drop in salinity during an eight day period coincided with dust input. It then becomes difficult to parse these two variables out as either may have contributed to growth at this site.

>>>>

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year [@dalmeidaModelSaharanDust1986]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico. These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria [@kelloggCharacterizationAerosolizedBacteria2004], fungi [@ramirez-camejoFungalDiversitySahara2022], virus-like particles [@griffinAfricanDesertDust2001], and minerals [@formentiChemicalCompositionMineral2003]. These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron [@millsIronPhosphorusColimit2004; @savoieNonseasaltSulfateNitrate1989; @grahamAtmosphericTransportPhosphorus1982]. This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups [@lenesSaharanDustPhosphatic2008; @westrichSaharanDustNutrients2016a ; @westrichSaharanDustNutrients2016a].

@@

Previous work has characterized *Vibrio* blooms in response to Saharan dust input in the oligotrophic setting of the Caribbean and subtropical Atlantic (Barbados and Florida Keys, respectively), with surface water concentrations of total *Vibrio* increasing by five to thirty times that found during non-dust conditions and returning to baseline levels within 24-48 hours. *Vibrio* composition within the larger microbial community also shifted following dust deposition, with initial levels of \<1.4% to a peak of 19.8% of the bacterial community [@westrichSaharanDustNutrients2016a]. A similar phenomenon was also observed in the surface waters of the tropical and open ocean mid-Atlantic with *Vibrio* populations increasing 1.5-fold in the mid-Atlantic following deposition [@westrichVibrioPopulationDynamics2018a] In a follow up study in the Florida Keys, episodic dust events during a daily time series promoted a succession of bacterial responses, with declines in *Prochlorococcus* coinciding with initial increases in bacteria belonging to the order *Vibrionales*, followed by subsequent shifts in response of different bacterial groups [@borchardtSaharanDustDeposition2020b]. To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients. It is unknown how comparable these findings are to coastal areas with higher ambient nutrient levels than these oligotrophic environments. The composition of desert dust can be complex and may have the potential to deliver critical resources that could be exploited by opportunistic microbes (like *Vibrio*) and elicit a growth response despite higher baseline nutrient levels. Given that human exposure is likely in these coastal waters, it is important to elucidate the relationship between dust input and microbial blooms in these higher nutrient environments.

@@

Sampling took place at three locations in Corpus Christi, TX: Blind Oso Bay, a residential canal system on Padre Island, and the Gulf. Sites were chosen based on differing baseline nutrient levels (defined here as the 2022 average from monthly observations). Blind Oso Bay is a shallow tributary that is popular amongst wadefisherman and kayakers. The Bay is often impacted by fluctuating salinities and high levels of nutrients (average:13.03 µM nitrate+nitrite, 4.44 µM orthophosphate) and chlorophyll due to its proximity to wastewater treatment discharge (Wetz, 2014).­ Additionally, the Bay has persistent issues with high levels of fecal indicator bacteria, placing it on the U.S. impaired waters list (Nicolau & Hill, 2014, Texas Commission on Environmental Quality, 2022). The residential canal system (average: 0.69µM nitrate+nitrite, 0.20 µM orthophosphate), and can also be impacted by high nutrients and chlorophyll from storm water runoff; salinity is primarily driven by precipitation. The Gulf site is characterized by constant salinities (\~36) and relatively low baseline nutrient levels (average: 0.19µM nitrate+nitrite and 0.36µM orthophosphate).

@@

This project works with three primary data sets. (1) Enumeration of Vibrio bacteria as copies/mL from qPCR (copies_master.csv), (2) Dust concentrations as aerosol optical thickness (AOT) derived from the Naval Research Lab (nrl_conc.csv), and (3) Environmental Parameters across the daily time series (ysi_2.csv). Below is a brief summary of how each data set was cleaned. More details can be found in the supplementary files, containing the code and comments describing what each line does. Please refer to the `project_README.md` file located in the NathanGreenslit-MADA-Project folder for details on the repository contents and instructions on reproducibility.\

@@

Quantitative PCR provides enumeration results as "cycle threshold or Cq" values. This depicts the cycle number in which enough of the DNA target was present to be amplified and thus detected. Lower Cq values correspond to higher target concentrations (as it takes less cycles to amplify), and vice verse for higher Cq values. The cleaning of this data set consisted of converting these Cq values into something that can be used to quantify Vibrio (copies of target per mL of seawater). To do so, I needed to take into account the amount of water sample that was concentrated on a filter, the total amount of DNA that was eluted following a DNA extraction, and the amount of DNA template added to each mix for qPCR (to name a few). Taking these into account, we are able to calculate our way from Cq values to copies of Vibrio per mL of seawater. qPCR was conducted in triplicate, so the last step was to take the average of the three replicates to have a final value per sample. Samples that were non-detect recieved a value of 0 copies/mL. A date column was also added for each sample.

@@

This figure depicts the summed dust concentration 24hr prior to sample collection (time points 13hr--> 7hr the next day) across the daily time series. The daily time series captured two dust events. One light event starting on the 13th of July, where AOD reached 0.2, and a much heavier dust event from the 16th to the 18th of July with AOD = 0.55-0.57. Dust AOD returned to lower concentrations following the time series.

@@

```{r, echo=FALSE, message=FALSE, fig.cap="Figure 3: Envonmental parameters across daily time series with color by site(Blind Oso = green, Canals = red, and the Gulf = Blue"}
knitr::include_graphics(here("results", "plots", "env.png"))
```

@@

Previous literature has shown that Vibrio respond 12-24 hours following dust deposition events [@westrichSaharanDustNutrients2016a]. Based on this, we need to look at lags between the two times series for dust data and growth data. To do this, we can set up a cross-correlation analysis, which allows us to examine the lag/lead relationship between the two variables.

CCF plots (found in supplemtary scripts) showed that BO has a significant lag relationship a value of -1. In other words, dust occurs, and one day later, we see a response in *Vibrio* gene copies per mL. C2 has a lag at -2, and RD at -1.

@@

#### Multi-varaiate Linear Regressions

@@

Multi-variate models that included dust, water chemistry, nutrients, and DOM (Model 5) resulted in the best performance, with mean RMSE values much lower than the Null Model.

@@

*Vibrio* gene copies/mL was found to significantly correlate with dust aerosol optical density at Blind Oso (r = 0.7, p = 0.01), Canals (r = 0.75, p = 0.005), and the Gulf (r = 0.67, p = 0.02). The introduction of dust constituents have the potential to serve as a temporary alleviation from the limitation of nutrients that are essential for bloom formation. *Vibrio* species have one of the fastest doubling times among bacteria [@gibsonDistributionBacterialDoubling2018 ; @eagonPSEUDOMONASNATRIEGENSMARINE1962], and are able to exploit introduced nutrients due their rapid motility, diverse genomic repertoire, and iron-chelating complexes [@ringgaardChemotaxisArraysVibrio2018; @okadaVibriosCommonlyPossess2005 ; @payneVibrioIronTransport2016]. Initially comprising a small proportion of the marine microbial communities (\<1%), *Vibrio* can come to quickly dominate, and thus have been classified as 'oppurtunitrophs'. The obsereved declines in *Vibrio* gene copies/mL following deposition may be attributed to top down control, such as viral lysis [@molina-quirozInteractionsVibrioPhages2023], or bottom-up controls due to an exhaustion of the resources provided by the dust [@psennerAreRapidChanges1992].

Similar to previous findings set in the Florida Keys and North Atlantic [@westrichVibrioPopulationDynamics2018a; @westrichSaharanDustNutrients2016a], peaks in growth were observed 24-48 hours following dust deposition. Comparing peak *Vibrio* gene copies/mL to initial levels on the 9th of July (chosen to ameliorate the possibility of growth following dust deposition prior to sampling period), the Gulf experienced a 33-fold increase in *Vibrio* gene copies/mL, Blind Oso a 2-fold increase, and the Canals a 5.6-fold increase.

@@

The Gulf exhibited a sharper increase in growth during the second, heavier dust event, while Blind Oso exhibited a smaller increase during the heavy dust event (Figure 2). This difference in response may be attributed to the nutrient dynamics of the sites. In marine waters, nutrients are typically limiting due to tight microbial recycling. Therefore the introduction of nutrients at such a site may result in rapid uptake and utilization. At Blind Oso, a priming effect may be occurring, where dust constituents resulting from a secondary flux may not be as essential for growth following an initial deposition event, where the constituents are reaching a maximum in how much they can support growth.

@@

While it can be proposed that the degree of microbial response is contingent on the environmental settings of the site, further work is needed to look at each specific environmental parameter and assess how it may impact *Vibrio* abundance. At Blind Oso for example, the drastic drop in salinity during an eight day period coincided with dust input. It then becomes difficult to parse these two variables out as either may have contributed to growth at this site.
